---
title: "PSYC259:Statistics in R"
author: "Amanda Mae Woodward"
date: "11/2/2020"
output: html_document
---

#Learning Outcomes:
By the end of today's class, students should be able to: 
- conduct common parametric analyses in R
- conduct a generalized linear model in R
- conduct common nonparametric analyses in R
- run multilevel models and mixed models in R

**Disclaimer:** Covering every type of analysis in R could be an entire course by itself. Today, we'll cover **some** analyes you can do. If there are additional analyses you'd like to cover, please let me know and I'm happy to upload supplemental code or cover it in a later class (there is flexibility in the last couple of weeks!). 

Additionally, we will **not** cover interpretations in depth in this class. The goal is to teach you how to use R to run the tests, and adding interpretations for each test could make this into several quarter long courses. I am happy to talk about interpretations in office hours, or you will learn about them in your statistics courses. 

```{r}

```

###Learning Outcome 1: Conduct common parametric analyses in R
We'll go through correlations, t-tests, regression, and ANOVA. We'll go through nonparametric tests (that may fit your data better) in the next section. 

**Pearson correlation**
We'll practice running correlations using the dataset above. To do this, we'll look at the correlation between Evaluations for includers and excluders in the dataset. 
cor(x, y)
```{r}

```
**Note:** It's great that we can see the correlation between these two measures, but we don't have any additional information, ie information related to significance.We can use another function, cor.test, to get information about significance.
cor.test(x,y)
```{r}

```
We can change whether we our conducting a one tailed or a two tailed test by including an additional argument "alternative." It defaults to a two tailed test, but we can specify a one tailed test in either direction (greater or less) 
```{r}

```
**Note:** There are several ways to handle missing data. We aren't going to cover them right now, but we'll talk about it more in later weeks.

**t-tests**
We can run a variety of t-tests using the same function t.test(). 

A one sample t test can be computed by specifying mu in the arguments. 
t.test(variable, mu)
```{r}

```
There are two ways we can use this function when we have two variables (independent or paired). The first is to type our x and y variables in as we did in the correlation function above. 
```{r}

```
OR
we can type them in as a formula in R. Formulas typically take the form y ~ x. To show you this example, I need to reformat our wide data to long data (using what we did last week!)

```{r}

```
This changes the t test arguments to: 
t.test(formula, data)
```{r}

```

If we think about the structure of our data, our data seem to be from a repeated measure. It's okay if you didn't infer that, it's my data and I made them to be this way. This is why including readme files for your data (whether it's publicly available or you're sharing it with collaborators) is really important! 

To account for the lack of dependence between observations, we'll run a paired samples t test. The code looks pretty similar to above, but we'll use an additional argument. 
```{r}

```

You'll notice that the top of the t-test output says "Welch's Two sample t-test." This R function automatically assumes that the variances of each group are unequal. If we wanted to run a traditional paired-samples t-test, we need to include another argument. 
```{r}

```

Finally, we some times run one tailed vs two tailed tests, just like we did with the correlations. 
```{r}

```

####Correlation and T-test practice
1. Open the mtcars dataset. Find the correlation between mpg and hp
2. Conduct a significance test to determine if displament and miles per gallon are correlated. 
3. Conduct a two-tailed t-test examining whether the average mpg differs by transmission (am). 
4. Conduct a one-tailed t-test examining whether the average displacement(disp) differs  engine shape (vs). Specifically, test whether straight engines result in higher displacements.

**Regression**
Back to the Exclusion Dataset we imported. The code for a linear regression is really similar (ie identical)  to what we used for t-tests.
lm(DV ~ IV, data)
```{r}

```
I tend to save my linear models because it allows me to do a few useful things:
Just like we used summary to get a summary of our data, we can use the same function to learn more about our models
```{r}

```
str() is a new(ish) function that allows us to learn about the structure of our model. We can use this to get specific pieces of information, or additional information that "underlies" our model (eg residuals and fitted values)
```{r}

```

**Multiple Regression**
We can include additional factors and interaction terms to our models: 

```{r}

```
The : can be used instead of + to include an interaction in your model
```{r}

```

Using * instead of + will include both the individual predictors AND their interactions 
```{r}

```

The class of our data and the way data are entered matter for regression models. 
let's consider age:
```{r}

```
Data don't really look continuous. We can change age to a factor. This will influence our output.  
```{r}

```

We may also need to change the reference level for factors. For instance, in this case, we care how exclusion changes children's evaluations. 
```{r}

```

**Anova**
There are several ways you can get Anova results in R. There are differences in the ways that they handle interactions, but they are used in the same way. 
```{r}

```
I typically use Anova from the car package, but there are some exceptions. We can talk about them as they come up. 

I'd like to be **very clear** the above models we ran with the long dataset are not accurate. There's nothing to account for the repeated measures design. It's important to keep these things in mind. Jake will talk more about ways to account for this structure.  

####Regression and ANOVA practice
1. Use the mtcars dataset and create a linear model that predicts mpg from cylinder (cyl) and displacement. Print the results
2. Create the same model, but include an interaction term. 
3. Run an ANOVA predicting hp from the transmission variable. 

###Learning Outcome #2: Conduct a generalized linear model in R
Back to the exclusion data again! Sometimes, our outcome variables aren't normally distributed. We can use generalized linear models in this case. The format is the same, but there is an additional argument "family."  

Let's do an example where our outcome variable, playChoice, is binary. We need to make a numeric variable containing 1's and 0's. 
```{r}

```
You can specify other distributions (negative binomials, poisson)

**Aside:** You may need to center variables as we get into more complex analyses. You can do this using the scale function. This computes a z score for each score. 
```{r}

```

####Learning Outcome # 2 Practice
1. Run a model that predicts a car's transmission from mpg. 

###Learning Outcome 3: Nonparametric analyses in R
Nonparametric analyses are run similarly to their parametric versions. In the interest of time, we'll talk about biserial correlations, spearman correlations, Wilcoxon sign rank tests, and binomial tests. 

**biserial correlations**
Biserial correlations involve a binary outcome and a continuous variable. To run one in R, we need the ltm package. 
```{r}

```
the function is biserial.cor(continuous, binary)
```{r}

```
Mathematically, this is the same as the pearson's version. 
```{r}

```

**spearman's rho**
We can calculate spearman's rho and kendall's tau the same way. We just need to use the "method" argument for cor.test 
```{r}

```

**Wilcoxon sign rank test**
This is the nonparametric version of the t-test. It has the same arguments. We'll do one as an example. 
wilcox.test
```{r}

```

**binomial tests**
We use binomial tests to determine if something happens more often than chance. The function is binom.test and it has the following arguments: 
binom.test(successes, totalScores, probability)

for instance, if we have 10 true/false statements, and get 6 right. Does this differ from chance? 
```{r}

```
This is a two-tailed test, but can also do one tailed by specifying the alternative. 

####Learning Outcome 3 practice: 
1. using the mtcars dataset, run a correlation to determine the relationship between engine shape (vs) and hp. What test did you run and why? 
2. Run a wilcoxon sign rank test to determine if cylinder and gears. 
3. Run a binomial test to determine if the number of cars with manual transmission differs from chance. (hint: use the ? feature to learn more about the dataset.)
